---
title: "Slana Telemetry results annotation"
output: word_document
date: "2026-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 6, fig.width = 8, dpi=300)
```

```{r, results='hide', message=FALSE, fig.keep='none'}
source("FDS_2025/R/1_Slana_telemetry_fromobjectives.R")
```



## Data import and handling

### Seasons
 * Spawning was defined as the location at the 2024-10-16 survey
 * Winter was defined as the location at the 2025-03-17 survey
 * Summer was defined as the location at the 2025-07-22 survey if an observation existed, and the 2025-06-11 survey otherwise.
 
### Inside/outside the Lake
...was pretty obvious

### Inside/outside the spawning area
Assignment is shown in the plots below.  These are some pretty rough plots generated by Leaflet and I can make some better maps if you'd like.

[insert zoomed out/zoomed in maps]

### Distance & homerange analyses in riverdist
...were conducted using the subset of data WITH FATE = ALIVE


\pagebreak

## Results tables

### Sequential distance

This gives the net (directional) distance traveled by each fish, between each sequential pair of surveys.

* Each row is an instrumented fish
* Each column is a pair of surveys (0 to 1, 1 to 2, etc)
* If a fish was not observed during one of each pair of surveys, the entry is recorded as NA
* Net upstream movement is recorded as positive, downstream movement is recorded as negative.

Here's a very simple plot of overall trends, or lack thereof:

```{r}
par(family="serif")
plotseq(distance_seq, ylab="River Distance (rkm)", xlab="Survey pair", main="Sequential Distance")
abline(h=0, lty=3)
```

\pagebreak

### Sequential direction

This gives the general direction (upstream vs. downstream) traveled by each fish, between each sequential pair of surveys.

* Each row is an instrumented fish
* Each column is a pair of surveys (0 to 1, 1 to 2, etc.)
* If a fish was not observed during one of each pair of surveys, the entry is recorded as NA

Here's a very simple plot of overall trends, or lack thereof:

```{r}
par(family="serif")
par(mar=par("mar")+c(-1,0,2,0))
mosaicplot(t(raw_dirtable), col=c(4, "grey80", 2), main="Sequential Direction", xlab="Survey pair", las=2)
```

\pagebreak

### Sequential distance by season

This gives the net (directional) distance traveled by each fish, between each pair of seasons

* Each row is an instrumented fish
* Each column is a pair of seasons
* If a fish was not observed during one of each pair of seasons, the entry is recorded as NA
* Net upstream movement is recorded as positive, downstream movement is recorded as negative.

Here's a very simple plot of overall trends, or lack thereof:

```{r}
par(family="serif")
boxplot(distance_seq_byseason, ylab="River Distance (rkm)", xlab="Survey pair", main="Sequential Distance")
abline(h=0, lty=3)
```


\pagebreak

### By Individual

This is a data table summarizing study-long results FOR EACH FISH.  Maybe not the most interesting, but could be used for further summaries or included as a data appendix. 

Columns include:

* *n_surveys*: Number of surveys in which the individual was observed, including tagging
* *homerange*: Observed minimum homerange for that individual
  * note: this is study-long, NOT the subset observed 2025-09-08.
* *totaldist*: Observed cumulative distance for that individual 
  * note: if an individual were observed on surveys 0, 1, and 3 (not 2), this would reflect the distance from 0 to 1 plus the distance from 1 to 3.
* *inside_lake* and *outside_lake*: Number of surveys in which the individual was observed inside vs outside the lake
* *p_lake*: Estimated proportion (SE) of surveys inside lake for that individual
* *inside_spawnarea* and *outside_spawnarea*: Number of surveys in which the individual was observed inside vs outside the spawning area
* *p_spawnarea*: Estimated proportion (SE) of surveys inside spawning area for that individual


\pagebreak

### Homerange by Survey

A table of homerange UP TO a given survey.

* Each column is a survey, each row is an individual fish
* Surveys in which an individual is not observed are recorded as NA.

```{r}
thedates <- sort(unique(ptdata$Date[ptdata$Survey>0]))
onthedate <- homerange_byindiv_bysurvey[,which(thedates=="2025-09-08")]
```


From the column associated with 2025-09-08, there were `r sum(!is.na(onthedate))` fish recorded, with a mean homerange of `r round(mean(onthedate, na.rm=TRUE),2)` (SE `r round(dsftools::se(onthedate, na.rm=TRUE),2)`) rkm, and a max and min of `r round(max(onthedate, na.rm=TRUE),2)` and `r round(min(onthedate, na.rm=TRUE),2)` rkm, respectively.

Plotting this two ways (first not incorporating missing observations, then imputing the value at the previous observation to make the plot fully cumulative) shows that the majority of movements making up an individual's homerange occured before the 2025-09-08 survey (survey `r which(thedates=="2025-09-08")`).


```{r, fig.height=4}
par(family="serif")
par(mar=par("mar")+c(-3.5,0,-1,0))
plotseq(homerange_byindiv_bysurvey, type="dotline", x="Survey", y="Homerange (rkm)", main="Cumulative homerange")
lines(colMeans(homerange_byindiv_bysurvey, na.rm=TRUE), lwd=3, lty=2)
```


```{r, fig.height=4}
par(family="serif")
par(mar=par("mar")+c(-3.5,0,-1,0))
plotseq(homerange_byindiv_bysurvey_fullycum, type="dotline", x="Survey", y="Homerange (rkm)", main="Cumulative homerange")
lines(colMeans(homerange_byindiv_bysurvey_fullycum, na.rm=TRUE), lwd=3, lty=2)
```







### Cumulative Distance by Survey

A table of cumulative distance UP TO a given survey.  If an individual were observed on surveys 0, 1, and 3 (not 2), this would reflect the distance from 0 to 1 plus the distance from 1 to 3.  Each column is a survey, each row is an individual fish.


```{r, fig.height=4}
par(family="serif")
par(mar=par("mar")+c(-3.5,0,-1,0))
plotseq(totaldist_byindiv_bysurvey_fullycum, type="dotline", x="Survey", y="Homerange (rkm)", main="Cumulative homerange")
lines(colMeans(totaldist_byindiv_bysurvey_fullycum, na.rm=TRUE), lwd=3, lty=2)
```
